---
title: "02_r4ds"
output: html_document
date: "2024-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(ggplot2)
library(tidyverse)
library(palmerpenguins)
library(ggthemes)


```


2  Workflow: basics

You can create new objects with the assignment operator <-

You can combine multiple elements into a vector with c()

And basic arithmetic on vectors is applied to every element of of the vector:


```{r}
primes <- c(2, 3, 5, 7, 11, 13)
primes * 2
primes - 1


```
All R statements where you create objects, assignment statements, have the same form:
object_name <- value
When reading that code, say “object name gets value” in your head.

You will make lots of assignments, and <- is a pain to type. You can save time with RStudio’s keyboard shortcut: Alt + - (the minus sign). Notice that RStudio automatically surrounds <- with spaces, which is a good code formatting practice. Code can be miserable to read on a good day, so giveyoureyesabreak and use spaces.


 2.2 Comments
 
 R will ignore any text after # for that line. This allows you to write comments, text that is ignored by R but read by other humans. We’ll sometimes include comments in examples explaining what’s happening with the code.
 
 With short pieces of code like this, leaving a comment for every single line of code might not be necessary. But as the code you’re writing gets more complex, comments can save you (and your collaborators) a lot of time figuring out what was done in the code.

Use comments to explain the why of your code, not the how or the what. The what and how of your code are always possible to figure out, even if it might be tedious, by carefully reading it. If you describe every step in the comments, and then change the code, you will have to remember to update the comments as well or it will be confusing when you return to your code in the future.


 2.3 What’s in a name?
 
 Object names must start with a letter and can only contain letters, numbers, _, and .. You want your object names to be descriptive, so you’ll need to adopt a convention for multiple words. We recommend snake_case, where you separate lowercase words with _.
 
i_use_snake_case
otherPeopleUseCamelCase
some.people.use.periods
And_aFew.People_RENOUNCEconvention


 2.4 Calling functions

R has a large collection of built-in functions that are called like this:

function_name(argument1 = value1, argument2 = value2, ...)

Let’s try using seq(), which makes regular sequences of numbers, and while we’re at it, learn more helpful features of RStudio. Type se and hit TAB. A popup shows you possible completions. Specify seq() by typing more (a q) to disambiguate or by using ↑/↓ arrows to select. Notice the floating tooltip that pops up, reminding you of the function’s arguments and purpose. If you want more help, press F1 to get all the details in the help tab in the lower right pane.

```{r}
seq(from =1, to = 10)
seq(1,10)


```

 2.5 Exercises
 
 
 Why does this code not work?
 
```{r}
my_variable <- 10
my_variable
```
 
 Tweak each of the following R commands so that they run correctly:
```{r}
library(tidyverse)

ggplot(data = mpg,
       mapping = aes(x=displ, y=hwy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```
 
 Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?
 
 Tools->keyboard shortcuts help
 
 Let’s revisit an exercise from the Section 1.6. Run the following lines of code. Which of the two plots is saved as mpg-plot.png? Why?
 
 
```{r}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```
 
 