---
title: "R Notebook"
output: html_notebook
---


4  Workflow: code style

Good coding style is like correct punctuation: you can manage without it, butitsuremakesthingseasiertoread. Even as a very new programmer, it’s a good idea to work on your code style. Using a consistent style makes it easier for others (including future-you!) to read your work and is particularly important if you need to get help from someone else. This chapter will introduce the most important points of the tidyverse style guide, which is used throughout this book.

Styling your code will feel a bit tedious to start with, but if you practice it, it will soon become second nature. Additionally, there are some great tools to quickly restyle existing code, like the styler package by Lorenz Walthert. Once you’ve installed it with install.packages("styler"), an easy way to use it is via RStudio’s command palette. The command palette lets you use any built-in RStudio command and many addins provided by packages. Open the palette by pressing Cmd/Ctrl + Shift + P, then type “styler” to see all the shortcuts offered by styler. Figure 4.1 shows the results.


```{r}

library(tidyverse)
library(nycflights13)

```

 4.1 Names

We talked briefly about names in Section 2.3. Remember that variable names (those created by <- and those created by mutate()) should use only lowercase letters, numbers, and _. Use _ to separate words within a name.

# Strive for:
short_flights <- flights |> filter(air_time < 60)

# Avoid:
SHORTFLIGHTS <- flights |> filter(air_time < 60)


 4.2 Spaces

Put spaces on either side of mathematical operators apart from ^ (i.e. +, -, ==, <, …), and around the assignment operator (<-).

# Strive for
z <- (a + b)^2 / d

# Avoid
z<-( a + b ) ^ 2/d

Don’t put spaces inside or outside parentheses for regular function calls. Always put a space after a comma, just like in standard English.

# Strive for
mean(x, na.rm = TRUE)

# Avoid
mean (x ,na.rm=TRUE)

It’s OK to add extra spaces if it improves alignment. For example, if you’re creating multiple variables in mutate(), you might want to add spaces so that all the = line up.1 This makes it easier to skim the code.

flights |> 
  mutate(
    speed      = distance / air_time,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %%  100
  )
  
  
4.3 Pipes

|> should always have a space before it and should typically be the last thing on a line. This makes it easier to add new steps, rearrange existing steps, modify elements within a step, and get a 10,000 ft view by skimming the verbs on the left-hand side.

# Strive for 
flights |>  
  filter(!is.na(arr_delay), !is.na(tailnum)) |> 
  count(dest)

# Avoid
flights|>filter(!is.na(arr_delay), !is.na(tailnum))|>count(dest)

After the first step of the pipeline, indent each line by two spaces. RStudio will automatically put the spaces in for you after a line break following a |> . If you’re putting each argument on its own line, indent by an extra two spaces. Make sure ) is on its own line, and un-indented to match the horizontal position of the function name.

# Strive for 
flights |>  
  group_by(tailnum) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights|>
  group_by(tailnum) |> 
  summarize(
             delay = mean(arr_delay, na.rm = TRUE), 
             n = n()
           )

# Avoid
flights|>
  group_by(tailnum) |> 
  summarize(
  delay = mean(arr_delay, na.rm = TRUE), 
  n = n()
  )
  
  
It’s OK to shirk some of these rules if your pipeline fits easily on one line. But in our collective experience, it’s common for short snippets to grow longer, so you’ll usually save time in the long run by starting with all the vertical space you need.

# This fits compactly on one line
df |> mutate(y = x + 1)

# While this takes up 4x as many lines, it's easily extended to 
# more variables and more steps in the future
df |> 
  mutate(
    y = x + 1
  )


4.4 ggplot2

The same basic rules that apply to the pipe also apply to ggplot2; just treat + the same way as |>.

flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = month, y = delay)) +
  geom_point() + 
  geom_line()


Again, if you can’t fit all of the arguments to a function on to a single line, put each argument on its own line:

flights |> 
  group_by(dest) |> 
  summarize(
    distance = mean(distance),
    speed = mean(distance / air_time, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = distance, y = speed)) +
  geom_smooth(
    method = "loess",
    span = 0.5,
    se = FALSE, 
    color = "white", 
    linewidth = 4
  ) +
  geom_point()

Watch for the transition from |> to +. We wish this transition wasn’t necessary, but unfortunately, ggplot2 was written before the pipe was discovered.


 4.5 Sectioning comments

As your scripts get longer, you can use sectioning comments to break up your file into manageable pieces:

# Load data --------------------------------------

# Plot data --------------------------------------

RStudio provides a keyboard shortcut to create these headers (Cmd/Ctrl + Shift + R), and will display them in the code navigation drop-down at the bottom-left of the editor, as shown in Figure 4.2.


 4.6 Exercises

    Restyle the following pipelines following the guidelines above.

    flights|>filter(dest=="IAH")|>group_by(year,month,day)|>summarize(n=n(),
    delay=mean(arr_delay,na.rm=TRUE))|>filter(n>10)

    flights|>filter(carrier=="UA",dest%in%c("IAH","HOU"),sched_dep_time>
    0900,sched_arr_time<2000)|>group_by(flight)|>summarize(delay=mean(
    arr_delay,na.rm=TRUE),cancelled=sum(is.na(arr_delay)),n=n())|>filter(n>10)
    
```{r exercise}

flights |>
  filter(dest=="IAH") |>
  group_by(year, month, day) |>
  summarize(n = n(), delay = mean(arr_delay, na.rm=TRUE)) |>
  filter(n > 10)

flights |>
  filter(
    carrier == "UA", 
    dest%in%c("IAH", "HOU"),
    sched_dep_time > 0900, 
    sched_arr_time < 2000
    ) |>
  group_by(flight) |>
  summarize(delay = mean(arr_delay,na.rm=TRUE), 
            cancelled=sum(is.na(arr_delay)), 
            n=n()
  ) |>
  filter(n > 10)



```

