---
title: "5  Data tidying"
output: html_notebook
---

```{r load libraries}

library(tidyverse)


```


5  Data tidying

#5.3 Lengthening data

The principles of tidy data might seem so obvious that you wonder if you’ll ever encounter a dataset that isn’t tidy. Unfortunately, however, most real data is untidy. There are two main reasons:

    Data is often organized to facilitate some goal other than analysis. For example, it’s common for data to be structured to make data entry, not analysis, easy.

    Most people aren’t familiar with the principles of tidy data, and it’s hard to derive them yourself unless you spend a lot of time working with data.

This means that most real analyses will require at least a little tidying. You’ll begin by figuring out what the underlying variables and observations are. Sometimes this is easy; other times you’ll need to consult with the people who originally generated the data. Next, you’ll pivot your data into a tidy form, with variables in the columns and observations in the rows.

tidyr provides two functions for pivoting data: pivot_longer() and pivot_wider(). We’ll first start with pivot_longer() because it’s the most common case. Let’s dive into some examples.

#5.3.1 Data in column names

The billboard dataset records the billboard rank of songs in the year 2000: In this dataset, each observation is a song. The first three columns (artist, track and date.entered) are variables that describe the song. Then we have 76 columns (wk1-wk76) that describe the rank of the song in each week1. Here, the column names are one variable (the week) and the cell values are another (the rank).

To tidy this data, we’ll use pivot_longer():

```{r pivot longer}

billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )

```

This data is now tidy, but we could make future computation a bit easier by converting values of week from character strings to numbers using mutate() and readr::parse_number(). parse_number() is a handy function that will extract the first number from a string, ignoring all other text.

```{r pivot longer, drop NA obversations}
billboard_longer <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(
    week = parse_number(week)
  )

billboard_longer
```

Now that we have all the week numbers in one variable and all the rank values in another, we’re in a good position to visualize how song ranks vary over time. The code is shown below and the result is in Figure 5.2. We can see that very few songs stay in the top 100 for more than 20 weeks.

```{r plot pivot longer}

billboard_longer |> 
  ggplot(aes(x = week, y = rank, group = track)) +
  geom_line(alpha = 0.25) +
  scale_y_reverse()

```

# 5.3.2 How does pivoting work?

Now that you’ve seen how we can use pivoting to reshape our data, let’s take a little time to gain some intuition about what pivoting does to the data. Let’s start with a very simple dataset to make it easier to see what’s happening. Suppose we have three patients with ids A, B, and C, and we take two blood pressure measurements on each patient. We’ll create the data with tribble(), a handy function for constructing small tibbles by hand:

```{r construct tibble by hand with tribble()}
df <- tribble(
  ~id, ~bp1, ~bp2,
  "A", 100, 120,
  "B", 140, 115,
  "C", 120, 125
)

```

We want our new dataset to have three variables: id (already exists), measurement (the column names), and value (the cell values). To achieve this, we need to pivot df longer:

```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```

# 5.3.3 Many variables in column names

A more challenging situation occurs when you have multiple pieces of information crammed into the column names, and you would like to store these in separate new variables. For example, take the who2 dataset, the source of table1 and friends that you saw above:


This dataset, collected by the World Health Organisation, records information about tuberculosis diagnoses. There are two columns that are already variables and are easy to interpret: country and year. They are followed by 56 columns like sp_m_014, ep_m_4554, and rel_m_3544. If you stare at these columns for long enough, you’ll notice there’s a pattern. Each column name is made up of three pieces separated by _. The first piece, sp/rel/ep, describes the method used for the diagnosis, the second piece, m/f is the gender (coded as a binary variable in this dataset), and the third piece, 014/1524/2534/3544/4554/5564/65 is the age range (014 represents 0-14, for example).

So in this case we have six pieces of information recorded in who2: the country and the year (already columns); the method of diagnosis, the gender category, and the age range category (contained in the other column names); and the count of patients in that category (cell values). To organize these six pieces of information in six separate columns, we use pivot_longer() with a vector of column names for names_to and instructors for splitting the original variable names into pieces for names_sep as well as a column name for values_to:

```{r pivot longer with many}

who2 |> 
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"),
    names_sep = "_",
    values_to = "count"
  )

```

An alternative to names_sep is names_pattern, which you can use to extract variables from more complicated naming scenarios, once you’ve learned about regular expressions in Chapter 15.


# 5.3.4 Data and variable names in the column headers

The next step up in complexity is when the column names include a mix of variable values and variable names. For example, take the household dataset:
This dataset contains data about five families, with the names and dates of birth of up to two children. The new challenge in this dataset is that the column names contain the names of two variables (dob, name) and the values of another (child, with values 1 or 2). To solve this problem we again need to supply a vector to names_to but this time we use the special ".value" sentinel; this isn’t the name of a variable but a unique value that tells pivot_longer() to do something different. This overrides the usual values_to argument to use the first component of the pivoted column name as a variable name in the output.

```{r}

household |> 
  pivot_longer(
    cols = !family,
    names_to = c(".value", "child"),
    names_sep = "_",
    values_drop_na = TRUE
  )

```

Figure 5.7 illustrates the basic idea with a simpler example. When you use ".value" in names_to, the column names in the input contribute to both values and variable names in the output.

# 5.4 Widening data

So far we’ve used pivot_longer() to solve the common class of problems where values have ended up in column names. Next we’ll pivot (HA HA) to pivot_wider(), which makes datasets wider by increasing columns and reducing rows and helps when one observation is spread across multiple rows. This seems to arise less commonly in the wild, but it does seem to crop up a lot when dealing with governmental data.

We’ll start by looking at cms_patient_experience, a dataset from the Centers of Medicare and Medicaid services that collects data about patient experiences:


The core unit being studied is an organization, but each organization is spread across six rows, with one row for each measurement taken in the survey organization. We can see the complete set of values for measure_cd and measure_title by using distinct()

Neither of these columns will make particularly great variable names: measure_cd doesn’t hint at the meaning of the variable and measure_title is a long sentence containing spaces. We’ll use measure_cd as the source for our new column names for now, but in a real analysis you might want to create your own variable names that are both short and meaningful.

pivot_wider() has the opposite interface to pivot_longer(): instead of choosing new column names, we need to provide the existing columns that define the values (values_from) and the column name (names_from)

```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```

The output doesn’t look quite right; we still seem to have multiple rows for each organization. That’s because, we also need to tell pivot_wider() which column or columns have values that uniquely identify each row; in this case those are the variables starting with "org":

```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

# 5.4.1 How does pivot_wider() work?
To understand how pivot_wider() works, let’s again start with a very simple dataset. This time we have two patients with ids A and B, we have three blood pressure measurements on patient A and two on patient B:
```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

We’ll take the values from the value column and the names from the measurement column:

```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

