---
title: "8 Visualize - Layers"
output: html_notebook
---

#  9.1 Introduction

In Chapter 1, you learned much more than just how to make scatterplots, bar charts, and boxplots. You learned a foundation that you can use to make any type of plot with ggplot2.

In this chapter, you’ll expand on that foundation as you learn about the layered grammar of graphics. We’ll start with a deeper dive into aesthetic mappings, geometric objects, and facets. Then, you will learn about statistical transformations ggplot2 makes under the hood when creating a plot. These transformations are used to calculate new values to plot, such as the heights of bars in a bar plot or medians in a box plot. You will also learn about position adjustments, which modify how geoms are displayed in your plots. Finally, we’ll briefly introduce coordinate systems.

We will not cover every single function and option for each of these layers, but we will walk you through the most important and commonly used functionality provided by ggplot2 as well as introduce you to packages that extend ggplot2.

```{r load libraries}

library(tidyverse)

```

#  9.2 Aesthetic mappings


#  9.2.1 Exercises

Create a scatterplot of hwy vs. displ where the points are pink filled in triangles.

```{r}

mpg |> 
  ggplot(aes(x = hwy, y = displ))+
  geom_point(shape = 17, color = "pink")

```
Why did the following code not result in a plot with blue points?

Because it's inside aes()!
```{r}


ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, color = "blue"))
```
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, stroke = 9), color = "blue")

```

What happens if you map an aesthetic to something other than a variable name, like aes(color = displ < 5)? Note, you’ll also need to specify x and y.

```{r}

mpg |> 
  ggplot(aes(x = displ, y = hwy, color = displ <5))+
  geom_point()

```

#  9.3 Geometric objects

Every geom function in ggplot2 takes a mapping argument, either defined locally in the geom layer or globally in the ggplot() layer. However, not every aesthetic works with every geom. You could set the shape of a point, but you couldn’t set the “shape” of a line. If you try, ggplot2 will silently ignore that aesthetic mapping. On the other hand, you could set the linetype of a line. geom_smooth() will draw a different line, with a different linetype, for each unique value of the variable that you map to linetype.

```{r}

# Left
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) + 
  geom_smooth()

# Right
ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) + 
  geom_smooth()

```

Here, geom_smooth() separates the cars into three lines based on their drv value, which describes a car’s drive train. One line describes all of the points that have a 4 value, one line describes all of the points that have an f value, and one line describes all of the points that have an r value. Here, 4 stands for four-wheel drive, f for front-wheel drive, and r for rear-wheel drive.

If this sounds strange, we can make it clearer by overlaying the lines on top of the raw data and then coloring everything according to drv.

```{r}

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) + 
  geom_point() +
  geom_smooth(aes(linetype = drv))

```

Many geoms, like geom_smooth(), use a single geometric object to display multiple rows of data. For these geoms, you can set the group aesthetic to a categorical variable to draw multiple objects. ggplot2 will draw a separate object for each unique value of the grouping variable. In practice, ggplot2 will automatically group the data for these geoms whenever you map an aesthetic to a discrete variable (as in the linetype example). It is convenient to rely on this feature because the group aesthetic by itself does not add a legend or distinguishing features to the geoms.

Geoms are the fundamental building blocks of ggplot2. You can completely transform the look of your plot by changing its geom, and different geoms can reveal different features of your data. For example, the histogram and density plot below reveal that the distribution of highway mileage is bimodal and right skewed while the boxplot reveals two potential outliers.

```{r}

# Left
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2)

# Middle
ggplot(mpg, aes(x = hwy)) +
  geom_density()

# Right
ggplot(mpg, aes(x = hwy)) +
  geom_boxplot()

```

ggplot2 provides more than 40 geoms but these don’t cover all possible plots one could make. If you need a different geom, we recommend looking into extension packages first to see if someone else has already implemented it (see https://exts.ggplot2.tidyverse.org/gallery/ for a sampling). For example, the ggridges package (https://wilkelab.org/ggridges) is useful for making ridgeline plots, which can be useful for visualizing the density of a numerical variable for different levels of a categorical variable. In the following plot not only did we use a new geom (geom_density_ridges()), but we have also mapped the same variable to multiple aesthetics (drv to y, fill, and color) as well as set an aesthetic (alpha = 0.5) to make the density curves transparent.

```{r}

library(ggridges)

ggplot(mpg, aes(x = hwy, y = drv, fill = drv, color = drv)) +
  geom_density_ridges(alpha = 0.5, show.legend = FALSE)

```
#  9.3.1 Exercises

What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

```{r}

mpg |> 
  ggplot(aes(x = displ, y = hwy))+
  geom_line()

mpg |> 
  ggplot(aes(x = displ, y = hwy))+
  geom_area()

mpg |> 
  ggplot(aes(x = displ))+
  geom_boxplot()

mpg |> 
  ggplot(aes(x = displ))+
  geom_histogram(binwidth = .25)

```

Earlier in this chapter we used show.legend without explaining it:

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv), show.legend = FALSE)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv))

```

What does show.legend = FALSE do here? What happens if you remove it? Why do you think we used it earlier?

What does the se argument to geom_smooth() do? - displays confidence interval, TRUE by default

Recreate the R code necessary to generate the following graphs. Note that wherever a categorical variable is used in the plot, it’s drv.

```{r}

mpg |> 
  ggplot(aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth(se = FALSE)

mpg |> 
  ggplot(aes(x = displ, y = hwy))+
  geom_point()+
  geom_smooth(aes(group = drv), se = FALSE)

mpg |> 
  ggplot(aes(x = displ, y = hwy, color = drv))+
  geom_point()+
  geom_smooth(se = FALSE)

mpg |> 
  ggplot(aes(x = displ, y = hwy))+
  geom_point(aes(color = drv))+
  geom_smooth(se = FALSE)

mpg |> 
  ggplot(aes(x = displ, y = hwy))+
  geom_point(aes(color = drv))+
  geom_smooth(aes(linetype = drv), se = FALSE)

mpg |> 
  ggplot(aes(x = displ, y = hwy, fill = drv))+
  geom_point(shape = 21, color = "white", stroke = 2)

```

#  9.4 Facets

In Chapter 1 you learned about faceting with facet_wrap(), which splits a plot into subplots that each display one subset of the data based on a categorical variable.

To facet your plot with the combination of two variables, switch from facet_wrap() to facet_grid(). The first argument of facet_grid() is also a formula, but now it’s a double sided formula: rows ~ cols.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl)
```

By default each of the facets share the same scale and range for x and y axes. This is useful when you want to compare data across facets but it can be limiting when you want to visualize the relationship within each facet better. Setting the scales argument in a faceting function to "free_x" will allow for different scales of x-axis across columns, "free_y" will allow for different scales on y-axis across rows, and "free" will allow both.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl, scales = "free")
```

#  9.4.1 Exercises

What happens if you facet on a continuous variable?

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cty)

```
What do the empty cells in the plot above with facet_grid(drv ~ cyl) mean? Run the following code. How do they relate to the resulting plot?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = drv, y = cyl))

```

What plots does the following code make? What does . do?

```{r}

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)

```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ cyl, nrow = 2)
```
What are the advantages to using faceting instead of the color aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?


Which of the following plots makes it easier to compare engine size (displ) across cars with different drive trains? What does this say about when to place a faceting variable across rows or columns?

```{r}

ggplot(mpg, aes(x = displ)) + 
  geom_histogram() + 
  facet_grid(drv ~ .)

ggplot(mpg, aes(x = displ)) + 
  geom_histogram() +
  facet_grid(. ~ drv)

```

Recreate the following plot using facet_wrap() instead of facet_grid(). How do the positions of the facet labels change?

```{r}

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point()+
  facet_wrap(~drv, ncol = 1)

```

#  9.5 Statistical transformations

Consider a basic bar chart, drawn with geom_bar() or geom_col(). The following chart displays the total number of diamonds in the diamonds dataset, grouped by cut. The diamonds dataset is in the ggplot2 package and contains information on ~54,000 diamonds, including the price, carat, color, clarity, and cut of each diamond. The chart shows that more diamonds are available with high quality cuts than with low quality cuts.

On the x-axis, the chart displays cut, a variable from diamonds. On the y-axis, it displays count, but count is not a variable in diamonds! Where does count come from? Many graphs, like scatterplots, plot the raw values of your dataset. Other graphs, like bar charts, calculate new values to plot:

    Bar charts, histograms, and frequency polygons bin your data and then plot bin counts, the number of points that fall in each bin.

    Smoothers fit a model to your data and then plot predictions from the model.

    Boxplots compute the five-number summary of the distribution and then display that summary as a specially formatted box.

The algorithm used to calculate new values for a graph is called a stat, short for statistical transformation. Figure 9.2 shows how this process works with geom_bar().

You can learn which stat a geom uses by inspecting the default value for the stat argument. For example, ?geom_bar shows that the default value for stat is “count”, which means that geom_bar() uses stat_count(). stat_count() is documented on the same page as geom_bar(). If you scroll down, the section called “Computed variables” explains that it computes two new variables: count and prop.

Every geom has a default stat; and every stat has a default geom. This means that you can typically use geoms without worrying about the underlying statistical transformation. However, there are three reasons why you might need to use a stat explicitly:

    You might want to override the default stat. In the code below, we change the stat of geom_bar() from count (the default) to identity. This lets us map the height of the bars to the raw values of a y variable.
    You might want to override the default mapping from transformed variables to aesthetics. For example, you might want to display a bar chart of proportions, rather than counts:
    You might want to draw greater attention to the statistical transformation in your code. For example, you might use stat_summary(), which summarizes the y values for each unique x value, to draw attention to the summary that you’re computing:
```{r}
ggplot(diamonds) + 
  stat_summary(
    aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```
ggplot2 provides more than 20 stats for you to use. Each stat is a function, so you can get help in the usual way, e.g., ?stat_bin.
    
#  9.5.1 Exercises

What is the default geom associated with stat_summary()? How could you rewrite the previous plot to use that geom function instead of the stat function?


```{r}
ggplot(diamonds) + 
  geom_pointrange(
    aes(x = cut, y = depth),
    stat = "summary",
    fun.min = min,
    fun.max = max,
    fun = median
    )
```
What does geom_col() do? How is it different from geom_bar()?

```{r}

diamonds |> 
  ggplot(aes(x = cut, y = depth))+
  geom_col()

diamonds |> 
  ggplot(aes(x = cut, y = depth))+
  geom_bar()


```

Most geoms and stats come in pairs that are almost always used in concert. Make a list of all the pairs. What do they have in common? (Hint: Read through the documentation.)

- geom_bar, stat_count
- geom_bin, stat_bin,
- geom_boxplot, stat_boxplot,
- geom_contour, stat_contour,
- geom_count, stat_sum,
- geom_density, stat_density,
-  geom_function() stat_function() 
- geom_quantile() stat_quantile() 
- geom_smooth() stat_smooth() 
- 
What variables does stat_smooth() compute? What arguments control its behavior?

Aids the eye in seeing patterns in the presence of overplotting. geom_smooth() and stat_smooth() are effectively aliases: they both use the same arguments. Use stat_smooth() if you want to display the results with a non-standard geom.

method, formula, se, span, fullrange, orientation


In our proportion bar chart, we needed to set group = 1. Why? In other words, what is the problem with these two graphs?

```{r}

ggplot(diamonds, aes(x = cut, y = after_stat(prop))) + 
  geom_bar(aes(group = 1))
ggplot(diamonds, aes(x = cut, fill = color, y = after_stat(prop))) + 
  geom_bar()

```

#  9.6 Position adjustments

There’s one more piece of magic associated with bar charts. You can color a bar chart using either the color aesthetic, or, more usefully, the fill aesthetic:

```{r}
# Left
ggplot(mpg, aes(x = drv, color = drv)) + 
  geom_bar()

# Right
ggplot(mpg, aes(x = drv, fill = drv)) + 
  geom_bar()
```
Note what happens if you map the fill aesthetic to another variable, like class: the bars are automatically stacked. Each colored rectangle represents a combination of drv and class.

```{r}
ggplot(mpg, aes(x = drv, fill = class)) + 
  geom_bar()
```

The stacking is performed automatically using the position adjustment specified by the position argument. If you don’t want a stacked bar chart, you can use one of three other options: "identity", "dodge" or "fill".

position = "identity" will place each object exactly where it falls in the context of the graph. This is not very useful for bars, because it overlaps them. To see that overlapping we either need to make the bars slightly transparent by setting alpha to a small value, or completely transparent by setting fill = NA.

```{r}
# Left
ggplot(mpg, aes(x = drv, fill = class)) + 
  geom_bar(alpha = 1/5, position = "identity")

# Right
ggplot(mpg, aes(x = drv, color = class)) + 
  geom_bar(fill = NA, position = "identity")
```

The identity position adjustment is more useful for 2d geoms, like points, where it is the default.

position = "fill" works like stacking, but makes each set of stacked bars the same height. This makes it easier to compare proportions across groups.

position = "dodge" places overlapping objects directly beside one another. This makes it easier to compare individual values.

```{r}
# Left
ggplot(mpg, aes(x = drv, fill = class)) + 
  geom_bar(position = "fill")

# Right
ggplot(mpg, aes(x = drv, fill = class)) + 
  geom_bar(position = "dodge")
```

There’s one other type of adjustment that’s not useful for bar charts, but can be very useful for scatterplots. Recall our first scatterplot. Did you notice that the plot displays only 126 points, even though there are 234 observations in the dataset?

The underlying values of hwy and displ are rounded so the points appear on a grid and many points overlap each other. This problem is known as overplotting. This arrangement makes it difficult to see the distribution of the data. Are the data points spread equally throughout the graph, or is there one special combination of hwy and displ that contains 109 values?

You can avoid this gridding by setting the position adjustment to “jitter”. position = "jitter" adds a small amount of random noise to each point. This spreads the points out because no two points are likely to receive the same amount of random noise.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter")

```
Adding randomness seems like a strange way to improve your plot, but while it makes your graph less accurate at small scales, it makes your graph more revealing at large scales. Because this is such a useful operation, ggplot2 comes with a shorthand for geom_point(position = "jitter"): geom_jitter().

To learn more about a position adjustment, look up the help page associated with each adjustment: ?position_dodge, ?position_fill, ?position_identity, ?position_jitter, and ?position_stack.

#  9.6.1 Exercises

What is the problem with the following plot? How could you improve it?

```{r}
ggplot(mpg, aes(x = cty, y = hwy)) + 
  geom_point()

mpg |> 
  ggplot(aes(x = cty, y = hwy)) +
  geom_jitter()
```

What, if anything, is the difference between the two plots? Why?

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(position = "identity", alpha = 0.3)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter()

```

What parameters to geom_jitter() control the amount of jittering?

width, height: Amount of vertical and horizontal jitter. The jitter is added in both positive and negative directions, so the total spread is twice the value specified here.

If omitted, defaults to 40% of the resolution of the data: this means the jitter values will occupy 80% of the implied bins. Categorical data is aligned on the integers, so a width or height of 0.5 will spread the data so it's not possible to see the distinction between the categories.



Compare and contrast geom_jitter() with geom_count().

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter() +
  geom_smooth()


ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_count() +
  geom_smooth()

```

What’s the default position adjustment for geom_boxplot()? Create a visualization of the mpg dataset that demonstrates it.

```{r}

mpg |> 
  ggplot(aes(class, hwy)) +
  geom_boxplot()

mpg |> 
  ggplot(aes(class, hwy)) +
  geom_boxplot(position = "dodge")

```

#  9.7 Coordinate systems

Coordinate systems are probably the most complicated part of ggplot2. The default coordinate system is the Cartesian coordinate system where the x and y positions act independently to determine the location of each point. There are two other coordinate systems that are occasionally helpful.

    coord_quickmap() sets the aspect ratio correctly for geographic maps. This is very important if you’re plotting spatial data with ggplot2. We don’t have the space to discuss maps in this book, but you can learn more in the Maps chapter of ggplot2: Elegant graphics for data analysis.
    
```{r}
nz <- map_data("nz")

ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black")

ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()
```
coord_polar() uses polar coordinates. Polar coordinates reveal an interesting connection between a bar chart and a Coxcomb chart.

```{r}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = clarity, fill = clarity), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1)

bar + coord_flip()
bar + coord_polar()
```
#  9.7.1 Exercises

Turn a stacked bar chart into a pie chart using coord_polar().

```{r}

mpg |> 
  filter(drv == "4") |> 
  ggplot(aes(x = class, fill = class)) +
  geom_bar(position = "fill")

mpg |> 
  filter(drv == "4") |> 
  ggplot(aes(x = drv, fill = class)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y")

```

What’s the difference between coord_quickmap() and coord_map()?

```{r}

nz <- map_data("nz")

ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black")

ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()

ggplot(nz, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_map()
```

What does the following plot tell you about the relationship between city and highway mpg? Why is coord_fixed() important? What does geom_abline() do?

```{r}

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```
geom_abline(): slope and intercept

this abline appears to be y=x

coord_fixed() ensures ratio of y/x remains constant when plot is resized

#  9.8 The layered grammar of graphics

We can expand on the graphing template you learned in Section 1.3 by adding position adjustments, stats, coordinate systems, and faceting:


```{r}


ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>

```

Our new template takes seven parameters, the bracketed words that appear in the template. In practice, you rarely need to supply all seven parameters to make a graph because ggplot2 will provide useful defaults for everything except the data, the mappings, and the geom function.

The seven parameters in the template compose the grammar of graphics, a formal system for building plots. The grammar of graphics is based on the insight that you can uniquely describe any plot as a combination of a dataset, a geom, a set of mappings, a stat, a position adjustment, a coordinate system, a faceting scheme, and a theme.





